---
title: "Updating ACSN maps"
author: "Paul Deveau"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
      self_contained: no
      fig_width: 7
      fig_height: 7
vignette: >
  %\VignetteIndexEntry{ACSN Map update}
  %\VignetteEngine{knitr::rmarkdown}
  %\usepackage[utf8]{inputenc}
---


# Updating ACSN maps walkthrough

## Rationale:

"I would like to change the default maps that are computed in ACSNMineR" or "ACSN maps have a new version, and I would like to update it before the package release." 

## A bit of storage:
ACSN maps are stored as a list of dataframes generated by:
````{r format, eval = FALSE}
  format_from_gmt 
```` 

They were exported using the R function "save" in the package folder "data/ACSN_maps.rda".
Changing this package and re-installing the modified ACSNMineR will thus change the maps on which enrichment is computed.

## Walkthrough
### 1. Generate your list of maps
<b> BEWARE </b> Once your map will be updated, the names that can be used as map argument in the enrichment function will be changed too!
In R the name of your map list should be ACSN_maps

### 2. Export it
If your working directory is the package main directory:
```{r export, eval = FALSE}
  save(list = ACSN_maps,file = "data/ACSN_maps.rda")

````

If you don't know where your working directory is, you can get the information with getwd() and change it with
```{r setwd, eval = FALSE}
  setwd("path/to/package")
````

### 3. Compile and install modified ACSNMineR

Now you only have to load devtools and install the package:
````{r devtools, eval = FALSE}
  library("devtools")
  install("/path/to/modified/ACSNMineR")
````

### 4. Load the new ACSNMineR and start you analysis
````{r library, eval = FALSE}
  library("ACSNMineR")

````
