---
output:
  html_document:
    highlight: tango
    toc: yes
  word_document:
    highlight: tango
---
# <a name="I">ACSN Enrichment</a>


##  <a name="IIA">Description </a>
ACSNEnrichment is an R package, freely available.

This package is designed for an easy analysis of gene maps (either user imported from gmt files or ACSN maps).
Its aim is to allow a statistical analysis of statistically enriched or depleted pathways from a user imported gene list, as well as a graphic representation of results.

This readme contains:

  [1. This description](#IIA)

  [2. Usage section](#IIB)

  [2.1. Pathway analysis](#IIIA)
  
  [2.1.1 Import gmt files](#IIIA1)
  
  [2.1.2 Perform analysis](#IIIA2)
  
  [2.2. Data vizualization](#IIIB)

  [2.2.1. Heatmaps](#IVA)

  [2.2.2. Barplots](#IVB)


##  <a name="IIB">Usage </a>
###  <a name="IIIA">Pathway analysis</a> 
______

#### <a name = "IIIA1"> Import gmt files </a>

Gmt files can be imported thanks to the format_from_gmt function. 
Let's use saved data from the package:

```{r}
# Retrieve path of the example gmt
file<-system.file("extdata", "cellcycle_short.gmt", package = "ACSNEnrichment")
# Then import it
gmt<-ACSNEnrichment::format_from_gmt(file)
knitr::kable(gmt[,1:10])
```

ACSN maps are built-in and can easily be accessed:
```{r}
# Available maps:
knitr::kable(t(names(ACSNEnrichment::ACSN_maps)))
#And accessing them:
knitr::kable(head(ACSNEnrichment::ACSN_maps$CellCycle[,1:10]), row.names = FALSE)

```

The gene set that was used for tests is the following:
```{r}
knitr::kable(ACSNEnrichment::genes_test)
```

#### <a name = "IIIA2"> Perform analysis </a>

Gene set enrichment for a single set can be performed by calling:
````{r}
Example<-ACSNEnrichment::enrichment(ACSNEnrichment::genes_test,
    min_module_size = 10, 
    threshold = 0.05,
    maps = list(cellcycle = ACSNEnrichment::ACSN_maps$CellCycle))
knitr::kable(Example,row.names = FALSE)
````

Where:

* genes_test is a character vector to test

* min_module_size is the minimal size of a module to be taken into account

* threshold is the maximal p-value that will be displayed in the results (all modules with p-values higher than threshold will be removed)

* maps is a list of maps -here we take the cell cycle map from ACSN-  imported through the format_from_gmt() function of the package

Gene set enrichment for multiple sets/cohorts can be performed by calling:

````{r}
Example<-ACSNEnrichment::multisample_enrichment(Genes_by_sample = list(set1 = ACSNEnrichment::genes_test[-1],
                                                              set2 = ACSNEnrichment::genes_test[-2]),
    maps = ACSNEnrichment::ACSN_maps$CellCycle,
    min_module_size = 10,
    cohort_threshold = FALSE)
knitr::kable(Example[[1]],row.names = FALSE)
knitr::kable(Example[[2]],row.names = FALSE)


````

Where:

* Genes_by_sample is a list of character vectors to test

* min_module_size is the minimal size of a module to be taken into account

* maps is a list of maps -here we take the cell cycle map from ACSN - imported through the format_from_gmt() function of the package


###  <a name="IIIB">Data visualization</a> 
______

Results from the enrichment analysis function can be transformed to images thanks to the represent enrichment function. Two different plot are available: heatmap and barplot.

####  <a name="IVA">Heatmaps</a>  
______

Heatmaps for single sample or multiple sample representing p-values can be easily generated thanks to the represent_enrichment function.

````{r}
ACSNEnrichment::represent_enrichment(enrichment = list(SampleA = ACSNEnrichment::enrichment_test[1:10,], 
    SampleB = ACSNEnrichment::enrichment_test[3:10,]),
    plot = "heatmap", 
    scale = "log",
    low = "steelblue" , high ="white",
    na.value = "grey")
````

Where:

* enrichment is the result from the enrichment or multisample_enrichment function

* scale can be set to either identity or log and will affect the gradient of colors

* low: the color for the low (significant) p-values 

* high: color for the high (less significant) p-values 

* na.value is the color in which tiles which have "NA" should appear


####  <a name="IVB">Barplots</a> 
______

A barplot can be achieved by using the following:
````{r}
ACSNEnrichment::represent_enrichment(enrichment = list(SampleA = ACSNEnrichment::enrichment_test[1:10,], 
    SampleB = ACSNEnrichment::enrichment_test[3:10,]),
    plot = "bar", 
    scale = "log",
    nrow = 1)

````
Where:

* enrichment is the result from the enrichment or multisample_enrichment function

* scale can be set to either identity or log and will affect the gradient of colors

* nrow is the number of rows that should be used to plot all barplots (default is 1)
