---
title: "ACSN description"
author: "Paul Deveau"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
      self_contained: no
      fig_width: 7
      fig_height: 7
vignette: >
  %\VignetteIndexEntry{ACSN description}
  %\VignetteEngine{knitr::rmarkdown}
  %\usepackage[utf8]{inputenc}

---
# <a name="I">ACSN Enrichment</a>

##  <a name="IIA">Description </a>
ACSNMineR is an R package, freely available.

ACSN stands for Atlas of Cancer Signaling Networks, and shows gene interaction in pathways relevant to cancer.

This package is designed for an easy analysis of gene maps (either user imported from gmt files or ACSN maps).
Its aim is to allow a statistical analysis of statistically enriched or depleted pathways from a user imported gene list, as well as a graphic representation of results.

This readme contains:

  [1. This description](#IIA)

  [2. Usage section](#IIB)

  [2.1. Pathway analysis](#IIIA)
  
  [2.1.1 Import gmt files](#IIIA1)
  
  [2.1.2 Perform analysis](#IIIA2)
  
  [2.2. Data vizualization](#IIIB)

  [2.2.1. Heatmaps](#IVA)

  [2.2.2. Barplots](#IVB)


##  <a name="IIB">Usage </a>
###  <a name="IIIA">Pathway analysis</a> 
______

#### <a name = "IIIA1"> Import gmt files </a>

Gmt files can be imported thanks to the format_from_gmt function. 
Let's use saved data from the package:

```{r gmt_example}
# Retrieve path of the example gmt
file<-system.file("extdata", "cellcycle_short.gmt", package = "ACSNMineR")
# Then import it
gmt<-ACSNMineR::format_from_gmt(file)
```
```{r gmt_display, echo = FALSE}
knitr::kable(gmt[,1:10])
```

ACSN maps are built-in and can easily be accessed through ACSNEnrcihment::ACSN_maps:
```{r gmt_map_code, eval = FALSE}
# Name of available maps:
names(ACSNMineR::ACSN_maps)

````
`````{r gmt_map_show, echo = FALSE}
knitr::kable(names(ACSNMineR::ACSN_maps))
````

````{r gmt_access_code, eval = FALSE}
#And accessing them:
ACSNMineR::ACSN_maps$CellCycle
````
````{r gmt_access_display, echo = FALSE}
knitr::kable(head(ACSNMineR::ACSN_maps$CellCycle[,1:10]), row.names = FALSE)
```

The gene set that was used for tests is the following:
````{r test_genes,eval = FALSE}
ACSNMineR::genes_test
````
```{r test_genes_show, echo = FALSE}
knitr::kable(ACSNMineR::genes_test)
```

#### <a name = "IIIA2"> Perform analysis </a>

Gene set enrichment for a single set can be performed by calling:
````{r Analysis_code}
Example<-ACSNMineR::enrichment(ACSNMineR::genes_test,
    min_module_size = 10, 
    threshold = 0.05,
    maps = list(cellcycle = ACSNMineR::ACSN_maps$CellCycle))
````
```{r Analysis_show, echo = FALSE}
knitr::kable(Example,row.names = FALSE)
````

Where:

* genes_test is a character vector to test

* min_module_size is the minimal size of a module to be taken into account

* threshold is the maximal p-value that will be displayed in the results (all modules with p-values higher than threshold will be removed)

* maps is a list of maps -here we take the cell cycle map from ACSN-  imported through the format_from_gmt() function of the package

Gene set enrichment for multiple sets/cohorts can be performed by calling:

````{r multi_analysis_code}
Example<-ACSNMineR::multisample_enrichment(Genes_by_sample = list(set1 = ACSNMineR::genes_test[-1],
                                                              set2 = ACSNMineR::genes_test[-2]),
    maps = ACSNMineR::ACSN_maps$CellCycle,
    min_module_size = 10,
    cohort_threshold = FALSE)
````
```{r multi_ana_code_1, eval = FALSE}
print(Example[[1]])
````
```{r multi_ana_show_1, echo = FALSE}
knitr::kable(Example[[1]],row.names = FALSE)
````
```{r multi_ana_code_2, eval = FALSE}
print(Example[[2]])
````
```{r multi_ana_show_2, echo = FALSE}
knitr::kable(Example[[2]],row.names = FALSE)

````

Where:

* Genes_by_sample is a list of character vectors to test

* min_module_size is the minimal size of a module to be taken into account

* maps is a list of maps -here we take the cell cycle map from ACSN - imported through the format_from_gmt() function of the package


###  <a name="IIIB">Data visualization</a> 
______

Results from the enrichment analysis function can be transformed to images thanks to the represent enrichment function. Two different plot are available: heatmap and barplot.

####  <a name="IVA">Heatmaps</a>  
______

Heatmaps for single sample or multiple sample representing p-values can be easily generated thanks to the represent_enrichment function.

````{r heatmap_plot}
ACSNMineR::represent_enrichment(enrichment = list(
    SampleA = ACSNMineR::enrichment_test[1:10,], 
    SampleB = ACSNMineR::enrichment_test[3:10,]),
    plot = "heatmap", 
    scale = "log",
    low = "steelblue" , high ="white",
    na.value = "grey")
````

Where:

* enrichment is the result from the enrichment or multisample_enrichment function

* scale can be set to either identity or log and will affect the gradient of colors

* low: the color for the low (significant) p-values 

* high: color for the high (less significant) p-values 

* na.value is the color in which tiles which have "NA" should appear


####  <a name="IVB">Barplots</a> 
______

A barplot can be achieved by using the following:
````{r barplot, warning = FALSE}
ACSNMineR::represent_enrichment(enrichment = list(
    SampleA = ACSNMineR::enrichment_test[1:10,], 
    SampleB = ACSNMineR::enrichment_test[3:10,]),
    plot = "bar", 
    scale = "log")

````

Where:

* enrichment is the result from the enrichment or multisample_enrichment function

* scale can be set to either identity or log and will affect the gradient of colors

* nrow is the number of rows that should be used to plot all barplots (default is 1)
